### 2013-02-15 doc/icml2013/intro.tex

% Discriminative models work better in practice Discriminative models
% with latent variables add expressive power (examples)
Discriminative latent-variable models,
which combine the high accuracy of discriminative models
with the compact expressiveness of latent-variable models,
have been widely applied to many tasks, including
object recognition \cite{quattoni04nips},
human action recognition \cite{wang09},
syntactic parsing \cite{petrov08discriminative},
and machine translation \cite{liang06discrimative}.
% http://www.cs.sfu.ca/~mori/research/papers/wang_cvpr09.pdf
%Slav Petrov, Discriminative Log-Linear Grammars with Latent Variables
However, parameter estimation in these models is difficult;
past approaches have relied exclusively on local optimization (EM, 
gradient descent) and are vulerable to local optima.
%Our goal is to develop consistent estimators which sidestep

% Goal of this paper
% This paper: first step in establishing provably correct estimation
% with simple mixture of linear regression
Our broad goal is to develop efficient provably consistent estimators for
discriminative latent-variable models.
In this paper, we provide a first step in this 
direction by proposing a new algorithm for a simple model,
\emph{mixture of linear regressions} (MLG) \cite{VieleTong2002}.
%mixture of linear regressions (\sectionref{sec:model}).

% Recent method of moments spectral methods sidestep local optima
Recently, method of moments estimators have been developed for
\emph{generative} latent-variable models, including
mixture models, HMMs \cite{AnandkumarHsuKakade2012},
Latent Dirichlet Allocation \cite{AnandkumarLDA},
and parsing models \cite{HsuKakadeLiang2012}.
The basic idea of this work is to express
the unknown model parameters as a tensor factorization
of the third-order moments of the model distribution, which
can be estimated from data.
The moments have a special symmetric structure
which permits the factorization to be computed efficiently using the robust
tensor power method \cite{AnandkumarGeHsu2012}.
%\cite{hsu09spectral}

@@

We use $x\tp{p}$ to represent the $p$-th order tensor
formed by taking the outer product of $x \in \Re^d$; i.e. $x\tp{p}_{i_1 i_2 \ldots
i_p} = x_{i_1} x_{i_2} \cdots x_{i_p}$. We will use
$\innerp{\cdot}{\cdot}$ to denote the generalized dot product between two $p$-th
order tensors: $\innerp{X}{Y} = \sum_{i_1, i_2, \ldots i_p} X_{i_1,
i_2, \ldots i_p} Y_{i_1, i_2, \ldots i_p}.$
For
$X \in (\Re^d)\tp{p}$ and $Y \in (\Re^d)\tp{(p+q)}$,% $p, q > 0$,
$\innerp{X}{Y} \in \Re^{d^q}$ is the projection of $X$ onto the first
$p$ modes of $Y$, i.e. $\innerp{X}{Y}_{i_1, \ldots, i_q}
= \sum_{j_1, \ldots, j_p} X_{j_1, \ldots, j_p} Y_{j_1, \ldots, j_p, i_1,
\ldots, i_q}$.
Let $[n] = \{ 1, \dots, n \}$ denote the first $n$ positive integers.
%Similar to how $x^T A$ is
%the projection of a vector $x$ onto the row space of a matrix $A$

%The expected moments of the parameters $\beta_1, \dots, \beta_k$ play an
%important role in our algorithm and we will use $M_p = \E[\beta\tp{p}]
%= \sum_{k=1}^{K} \pi_k \beta_k^{\otimes p}$ to denote them.

>>>>

% Discriminative models work better in practice Discriminative models
% with latent variables add expressive power (examples)
Discriminative latent-variable models,
which combine the high accuracy of discriminative models
with the compact expressiveness of latent-variable models,
have been widely applied to many tasks, including
object recognition \cite{quattoni04crf},
human action recognition \cite{wang09crf},
syntactic parsing \cite{petrov08discriminative},
and machine translation \cite{liang06discrimative}.
However, parameter estimation in these models is difficult;
past approaches have relied exclusively on local optimization (EM, 
gradient descent) and are vulerable to local optima.

% Goal of this paper
% This paper: first step in establishing provably correct estimation
% with simple mixture of linear regression
Our broad goal is to develop efficient provably consistent estimators for
discriminative latent-variable models.
In this paper, we provide a first step in this 
direction by proposing a new algorithm for a simple model,
\emph{mixture of linear regressions} (MLG) \cite{VieleTong2002}.

% Recent method of moments spectral methods sidestep local optima
Recently, method of moments estimators have been developed for
\emph{generative} latent-variable models, including
mixture models, HMMs \cite{AnandkumarHsuKakade2012},
Latent Dirichlet Allocation \cite{anandkumar12lda},
and parsing models \cite{hsu12identifiability}.
The basic idea of this work is to express
the unknown model parameters as a tensor factorization
of the third-order moments of the model distribution, which
can be estimated from data.
The moments have a special symmetric structure
which permits the factorization to be computed efficiently using the robust
tensor power method \cite{AnandkumarGeHsu2012}.

@@

We use $x\tp{p}$ to represent the $p$-th order tensor
formed by taking the outer product of $x \in \Re^d$; i.e. $x\tp{p}_{i_1 i_2 \ldots
i_p} = x_{i_1} x_{i_2} \cdots x_{i_p}$. We will use
$\innerp{\cdot}{\cdot}$ to denote the generalized dot product between two $p$-th
order tensors: $\innerp{X}{Y} = \sum_{i_1, i_2, \ldots i_p} X_{i_1,
i_2, \ldots i_p} Y_{i_1, i_2, \ldots i_p}.$
For a $p$-th order tensor $X \in (\Re^d)\tp{p}$, let $X_{(i)} \in \Re^{d \times d^{p-1}}$
be the mode-$i$ unfolding of $X$.
%For $X \in (\Re^d)\tp{p}$ and $Y \in (\Re^d)\tp{(p+q)}$,% $p, q > 0$,
%$\innerp{X}{Y} \in \Re^{d^q}$ is the projection of $X$ onto the first
%$p$ modes of $Y$, i.e. $\innerp{X}{Y}_{i_1, \ldots, i_q}
%= \sum_{j_1, \ldots, j_p} X_{j_1, \ldots, j_p} Y_{j_1, \ldots, j_p, i_1,
%\ldots, i_q}$.

% Norms
For a matrix $X$,
let $\|X\|_*$ denote the nuclear (trace) norm (sum of singular values),
let $\|X\|_F$ denote the Frobenius norm (square root of sum of squares of singular values),
let $\|X\|_\text{op}$ denote the operator norm (largest singular value).
For a rank-$K$ tensor $X$,
let $\|X\|_* = \frac{1}{p} \sum_{i=1}^p \|X_{(i)}\|_*$ denote
the average nuclear norm over all $p$ unfoldings,
and let $\|X\|_\text{op} = \frac{1}{p} \sum_{i=1}^p \|X_{(i)}\|_\text{op}$
denote the average operator norm over all $p$ unfoldings.
Let $\unfold(X)$ be the vectorized version of $X$.

Let $[n] = \{ 1, \dots, n \}$ denote the first $n$ positive integers.

### 2013-02-15 doc/icml2013/model.tex

The mixture of linear regressions model \citep{VieleTong2002} defines
a conditional distribution over a response $y \in \Re$
given covariates $x \in \Re^d$.
The generation of $y$ given $x$ involves two steps:
(i) draw the mixture component $h \in [K]$ according to mixture proportions
$\pi = (\pi_1, \dots, \pi_K)$;
and (ii) generate $y$ from a Gaussian centered at $\beta_h^\top x$ with
variance $\sigma^2$.
%Given $x$, the generative process for the response
%variables is,
\begin{eqnarray*}
  h &\sim& \mult(\pi), \\
  y \mid h &\sim& \normal{ \beta_{h}^T x}{ \sigma^2 }.
\end{eqnarray*}
%where $h$ is a latent variable corresponding to the mode.
The parameters of the model are $\theta = (\pi, B, \sigma^2)$,
where $\pi \in \Re^d$ are the mixture proportions,
$B = [\beta_1 \mid \dots \mid \beta_K] \in \Re^{d \times K}$
are the regression coefficients,
and $\sigma^2$ is the variance of the observation noise.
The learning problem is stated as follows:
given $N$ i.i.d. samples $(\xn{1}, \yn{1}), \dots, (\xn{N}, \yn{N})$
drawn from the model with some unknown parameters $\theta^*$,
return an estimate of the parameters $\hat\theta$.

%and would like to
%recover the regressors, $B = [\beta_1 \mid \beta_2 \mid \cdots \mid
%\beta_K]$, and the mixture probabilities, $\pi = (\pi_1, \pi_2, \dots,
%\pi_K)$.

%Note that we do not observe $h$. If we did, the optimal
%solution for $\beta_k$ would be the regressor found using least squares
%on the set of points, $\{(x_i, y_i) | h_i = k \}$.

% Typically people use EM, but the maximum marginal likelihood is
% non-convex.
The estimation problem is difficult primarily because the mixture
components $h$ are unobserved, rendering the marginal likelihood
a non-convex function of $\theta$.
Typically, the parameters are learned using
expectation maximization (EM) \cite{VieleTong2002},
which suffers from local optima
(we show this empirically in \sectionref{sec:evaluation}).
%We provide several instances where this leads
%to poor local optima in \sectionref{sec:evaluation}.

This model was used for XXX.

TODO: move to later
% Reference mixture of experts.
% Acknowledge that mixture proportions % can't depend on $x$
The mixture of linear regressions model is an instance of
mixture of experts \cite{jacobs91},
%wherein the components can be SVMs, GPs, regression trees and other estimators.
%In practice, it is
in which the mixture proportions are allowed to depend on $x$.
We have not considered these extensions in our treatment of the model.

>>>>

The mixture of linear regressions model \citep{VieleTong2002} defines
a conditional distribution over a response $y \in \Re$
given covariates $x \in \Re^d$.
The generation of $y$ given $x$ involves two steps:
(i) draw the mixture component $h \in [K]$ according to mixture proportions
$\pi = (\pi_1, \dots, \pi_K)$;
and (ii) generate $y$ from a Gaussian centered at $\beta_h^\top x$ with
variance $\sigma^2$.
\begin{eqnarray*}
  h &\sim& \mult(\pi), \\
  y \mid h &\sim& \normal{ \beta_{h}^T x}{ \sigma^2 }.
\end{eqnarray*}
The parameters of the model are $\theta = (\pi, B)$,
where $\pi \in \Re^d$ are the mixture proportions and
$B = [\beta_1 \mid \dots \mid \beta_K] \in \Re^{d \times K}$
are the regression coefficients.
We assume the variance of the observation noise $\sigma^2$ is known.
The learning problem is stated as follows:
given $n$ i.i.d. samples $(\xn{1}, \yn{1}), \dots, (\xn{n}, \yn{n})$
drawn from the model with some unknown parameters $\theta^*$,
return an estimate of the parameters $\hat\theta$.

% History of model
The mixture of linear regressions model has been applied
in the statistics literature for modeling music perception, where $x$ is the
actual tone and $y$ is the tone perceived by a musician \cite{VieleTong2002}.
% Reference mixture of experts.
% Acknowledge that mixture proportions % can't depend on $x$
The model is an instance of hierarchical mixture of experts
\cite{jacobs91experts}, in which the mixture proportions are allowed to depend
on $x$, known as a gating function.
This dependence allow the experts to be localized in input space,
providing more flexbility, but we do not consider this dependence in our model.

% Typically people use EM, but the maximum marginal likelihood is
% non-convex.
The estimation problem for mixture of linear regressions is difficult because
the mixture components $h$ are unobserved,
resulting in a non-convex log marginal likelihood.
The parameters are typically learned using
expectation maximization (EM) or Gibbs sampling \cite{VieleTong2002},
which suffers from local optima.
In the next section, we present a new algorithm
that sidesteps the local optima problem entirely.
%(we show this empirically in \sectionref{sec:evaluation}).
%We provide several instances where this leads
%to poor local optima in \sectionref{sec:evaluation}.

### 2013-02-15 doc/icml2013/regression.tex

In this section, we compute the restricted strong convexity constant
$\kappa(\opX)$ for our problem.

@@

%We begin our analysis by stating a generic theorem about convex
%optimization problem in \equationref{eq:regression}, 
\begin{lemma}[convergence of low-rank regression]
If restricted strong convexity holds
$$\frac{1}{2n} \| \opX( \Delta )\|_2^2 \ge \kappa(\opX) \|\Delta\|^2_F \quad \text{and} \quad
\lambda_n \ge \frac{|\opX^*(\eta_p)|}{n}.$$
Then the error of $\hat M_p$ is bounded as follows:
$\| \hat M_p - M_p^* \|_F \le \frac{\lambda_n \sqrt{k}}{\kappa(\opX)}$.
\end{lemma}
\begin{proof}
  \citet{Tomioka2011}, Theorem 1.

  This proof follows from convex properties \todo{elaborate}.
  \citet{NegahbanWainwright2009} prove this theorem for matrices, i.e.
  when $p = 2$. \citet{Tomioka2011} generalized it for arbitrary $p$.
\end{proof}

Going forward, we need to lower bound the restricted strong convexity
parameter $\kappa(\opX)$ and upper bound the adjoint operator
$\|\opX^*(\epsilon)\|_{2}^2$, which corresponds to a gradient update. We
will appeal to the random design framework that models the input $x$ as
random and show bounds that hold with high probability.

\begin{lemma}[restricted strong convexity]
Let $\opX$ be the linear operator previously defined. Then,
$$\kappa(\opX) \ge Y$$.
\end{lemma}
\begin{proof}
  \todo{Do I need to transform rewrite $M_p$ in terms of orthogonal components?}

  \begin{align*}
    \frac{\|\opX(M_p)\|_1}{N} &= \frac{1}{N}\sum_{i=1}^{N} |\innerp{x_i\tp{p}}{M_p}| \\
    &\ge \frac{1}{N}\sum_{i=1}^{N} \sigma_{min}( M_p ) \|x_i\|^p \\
    &\ge \frac{\kappa(M_p)}{\sqrt{D}} \frac{\sum_{i=1}^{N} \|x_i\|^p}{N} \| M_p \|_F.
  \end{align*}

  While $\kappa(M_p)$ depends on $M_p$ we can lower bound it by
  restricting the convex space to $\{ M_p | \kappa(M_p) \ge \kappa \}$.
  Further, using concentration bounds, we can lower bound $\|x_i\|^p$ as
  well. \todo{What is $\E[\|x_i\|^3]$? It is some quantity greater than
  zero}.

  Thus, 
  \begin{align*}
  \frac{ \|\opX(M_p)\|_2 }{2 N} 
  &\ge
  \frac{ \|\opX(M_p)\|_1 }{2 \sqrt{D}N} 
  &\ge \frac{\kappa}{2 D} (\E[\|x_i\|^p] - \frac{t}{N}) \| M_p \|_F.
  \end{align*}
  with high probability $1 - \delta$. \todo{The bound needs to be shown
  for $\| \opX(M_p) \|_2^2$.}

  In other words, $\kappa(\opX)
  = \frac{\kappa}{2 D} (\E[\|x_i\|^p] - \frac{t}{N})$ is a valid
  lower bound.
\end{proof}

@@

\begin{lemma}[bounded error]
  Let $\opX$ be the linear operator previously defined. Then,
  $$\opX^*(\epsilon) \le Y$$.
\end{lemma}
\begin{proof}
  \begin{align*}
    \frac{\opX^*(\epsilon)}{N} 
    &= \frac{1}{N}\sum_{i=1}^{N} \epsilon_i \innerp{x_i\tp{p}}{M_p} \\
    &= \frac{1}{N}\sum_{i=1}^{N} (\innerp{x_i\tp{p}}{M_p - \beta_i\tp{p}} + \varepsilon)  \innerp{x_i\tp{p}}{M_p}.
  \end{align*}
  Let $M_p - \beta_i\tp{p} = \tilde M_i$ and consider it to be a random
  variable.
  \begin{align*}
    \frac{\opX^*(\epsilon)}{N} 
    &= \frac{1}{N}\sum_{i=1}^{N} \innerp{\tilde M_i}{x_i\tp{2p}} + \varepsilon \innerp{M_p}{x_i\tp{p}} \\
    &\le \frac{1}{N}\sum_{i=1}^{N} \|\tilde M_i\|_{op} \|x_i\|^{2p} + \varepsilon \|M_p\|_{op}\|x_i\|^{p} \\
    &\le \|M_p\|_{op} ( \frac{1}{N}\sum_{i=1}^{N} \|x_i\|^{2p} + \varepsilon \|x_i\|^{p} ).
  \end{align*}

  We can probabilistically bound the above by considering the union
  bound of $\varepsilon$ and $\|x_i\|^p$.

  Finally,
  \begin{align*}
    \frac{\opX^*(\epsilon)}{N} 
    &\le \|M_p\|_{op} (\E[\|x_i\|^{2p}] + t/N + (t/N) (\E[ \|x_i\|^{p} ] + t/N) ),
  \end{align*}
  with high probability. Note that this does not converge to $0$ as $N
  \to \infty$ because of the variance of the selection probability.
\end{proof}

>>>>

In this section, we lower bound the restricted strong convexity constant
$\kappa(\opX)$.

\paragraph{Adjoint operator}

In this section, we upper bound the operator norm of the adjoint
$\|\opX(\eta_p)\|_\text{op}$.

@@

\begin{lemma}[\citet{Tomioka2011}, Theorem 1]
Suppose there exists a restricted strong convexity constant $\kappa(\opX)$ such that
$$\frac{1}{2n} \| \opX( \Delta )\|_2^2 \ge \kappa(\opX) \|\Delta\|^2_F \quad \text{and} \quad
\lambda_n \ge \frac{|\opX^*(\eta_p)|}{n}.$$
Then the error of $\hat M_p$ is bounded as follows:
$\| \hat M_p - M_p^* \|_F \le \frac{\lambda_n \sqrt{k}}{\kappa(\opX)}$.
\end{lemma}

Going forward, we need to lower bound the restricted strong convexity
parameter $\kappa(\opX)$ and upper bound the adjoint operator
$\|\opX^*(\epsilon)\|_{2}^2$.
%We
%will appeal to the random design framework that models the input $x$ as
%random and show bounds that hold with high probability.

\begin{lemma}[lower bound on restricted strong convexity]
Let $\Sigma_p \eqdef \E[\unfold(x\tp{p})\tp{2}]$.
With high probability at least $1-\delta$,
for some constant $c$,
$$\kappa(\opX) \ge c \sigma_\text{min}^p(\Sigma_p)$$,
where
%$\sigma_\text{min}^p(\Sigma_p) = \min_{M_p \in \sM_p : \|M_p\|_F \le 1} \unfold(M_p)^\top \Sigma_p \unfold(M_p)$.
$\sigma_\text{min}^p(\Sigma_p) = \min_{M_p \in \sM_p : \|M_p\|_F \le 1} \trace(\unfold(M_p)^{\otimes 2} \Sigma_p)$.
\end{lemma}
Note that $\sigma_\text{min}$ 
\begin{proof}
Expanding the definition of the observation operator:
$\frac{1}{2n} \|\opX(\Delta)\|_2^2
= \frac{1}{2n} \sum_{(x,y) \in \sD} \innerp{\Delta}{x\tp{p}}^2$.
Unfolding the tensor and letting $\hat\Sigma_p \eqdef \frac{1}{n} \sum_{(x,y) \in \sD} \unfold(x\tp{p})\tp{2}$,
we have 
$\frac{1}{2n} \|\opX(\Delta)\|_2^2
\ge \frac{1}{2} \trace(\unfold(\Delta)\tp{2} \hat\Sigma_p)
\ge \frac{1}{2} \|\Delta\|_F^2 \sigma_\text{min}^p(\hat\Sigma_p)$.

%  \begin{align*}
%    \frac{\|\opX(M_p)\|_1}{N} &= \frac{1}{N}\sum_{i=1}^{N} |\innerp{x_i\tp{p}}{M_p}| \\
%    &\ge \frac{1}{N}\sum_{i=1}^{N} \sigma_{min}( M_p ) \|x_i\|^p \\
%    &\ge \frac{\kappa(M_p)}{\sqrt{D}} \frac{\sum_{i=1}^{N} \|x_i\|^p}{N} \| M_p \|_F.
%  \end{align*}
%
%  While $\kappa(M_p)$ depends on $M_p$ we can lower bound it by
%  restricting the convex space to $\{ M_p | \kappa(M_p) \ge \kappa \}$.
%  Further, using concentration bounds, we can lower bound $\|x_i\|^p$ as
%  well. \todo{What is $\E[\|x_i\|^3]$? It is some quantity greater than
%  zero}.
%  
%  Thus, 
%  \begin{align*}
%  \frac{ \|\opX(M_p)\|_2 }{2 N} 
%  &\ge
%  \frac{ \|\opX(M_p)\|_1 }{2 \sqrt{D}N} 
%  &\ge \frac{\kappa}{2 D} (\E[\|x_i\|^p] - \frac{t}{N}) \| M_p \|_F.
%  \end{align*}
%  with high probability $1 - \delta$. \todo{The bound needs to be shown
%  for $\| \opX(M_p) \|_2^2$.}
%  
%  In other words, $\kappa(\opX)
%  = \frac{\kappa}{2 D} (\E[\|x_i\|^p] - \frac{t}{N})$ is a valid
%  lower bound.
\end{proof}

\begin{lemma}[bounded error]
  Let $\opX$ be the linear operator previously defined. Then,
  $$\|\opX^*(\epsilon)\|_\text{op} \le Y$$.
\end{lemma}
\begin{proof}
%  \begin{align*}
%    \frac{\opX^*(\epsilon)}{N} 
%    &= \frac{1}{N}\sum_{i=1}^{N} \epsilon_i \innerp{x_i\tp{p}}{M_p} \\
%    &= \frac{1}{N}\sum_{i=1}^{N} (\innerp{x_i\tp{p}}{M_p - \beta_i\tp{p}} + \varepsilon)  \innerp{x_i\tp{p}}{M_p}.
%  \end{align*}
%  Let $M_p - \beta_i\tp{p} = \tilde M_i$ and consider it to be a random
%  variable.
%  \begin{align*}
%    \frac{\opX^*(\epsilon)}{N} 
%    &= \frac{1}{N}\sum_{i=1}^{N} \innerp{\tilde M_i}{x_i\tp{2p}} + \varepsilon \innerp{M_p}{x_i\tp{p}} \\
%    &\le \frac{1}{N}\sum_{i=1}^{N} \|\tilde M_i\|_{op} \|x_i\|^{2p} + \varepsilon \|M_p\|_{op}\|x_i\|^{p} \\
%    &\le \|M_p\|_{op} ( \frac{1}{N}\sum_{i=1}^{N} \|x_i\|^{2p} + \varepsilon \|x_i\|^{p} ).
%  \end{align*}
%
%  We can probabilistically bound the above by considering the union
%  bound of $\varepsilon$ and $\|x_i\|^p$.
%
%
%  Finally,
%  \begin{align*}
%    \frac{\opX^*(\epsilon)}{N} 
%    &\le \|M_p\|_{op} (\E[\|x_i\|^{2p}] + t/N + (t/N) (\E[ \|x_i\|^{p} ] + t/N) ),
%  \end{align*}
%  with high probability. Note that this does not converge to $0$ as $N
%  \to \infty$ because of the variance of the selection probability.

@@

\end{proof}
